Forecasting short-term hourly Emergency Departement arrivals

### 2021-11-08 paper/paper.Rmd
Ivan has done his edits. New structure, some mistakes fixed and some additional explanations.

@morzuch2006forecasting used the Unobserved Components Model (UCM), by which each component of the time series is separately modeled as stochastic. Double-seasonal exponential smoothing and standard Holt-Winters are used to forecast ED arrival for an horizon of 168 hours. The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles. The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.

@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services. Hourly arrival data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future. Authors . They present the prediction interval accuracy of $50\%$ and $90\%$ intervals for the number of hourly arrivals under the Poisson assumptions. They show that the most important predictor is hour of the day and autocorrelation lag 1. However, these findings are limited to the short number of observations (only one year of historical data).

@schweigler2009forecasting conducted an investigation on whether using time series methods could accurately generate short-term forecasts of ED bed occupancy.  A year-long dataset of hourly ED bed occupancy was collected from three facilities. For each facility, the authors implemented an hourly historical average model, SARIMA model and sinusoidal model with autocorrelated error. In particular, the historical average model was based on the mean occupancy for each site for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept. They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal model outperformed the historical average (for example, at site 2, the two models improved by $33\%$ the 12-hour forecasts generated by historical average).

@kim2014predicting compared different univariate and multivariate time series forecasting techniques to forecast patient volume for a Hospital Medicine programme. The study adopted historical mean linear regression as benchmark, exponential smoothing, ARIMA, SARIMA, GARCH (generalized autoregressive conditional heteroskedasticity method), able to adjust changes in variance over time and vector autoregressive (VAR) method, able to incorporate data from different sources, to forecast for 4 hours, 24 hours. They used MAPE to report the forecast accuracy.
Each of the forecasting models outperformed the benchmark model. In particular, ARIMA model performed best.

@gijo2016sarima generated a time series model to forecast the daily and hourly call volume at all centre handling emergency ambulance services.  Since historical data showed seasonality, SARIMA models were investigated. Regarding the daily model, the authors generated a SARIMA model, which, however, resulted in the forecast error (MMSE) that significantly increased when the lead time exceeded 8 days. On the other hand, the SARIMA model proposed to forecast the log-calls an hourly basis. This model was found to fit well the model both for shorter and longer lead times.

@asheim2019real developed a Poisson time-series regression model with continuous weekly and yearly cyclic effects to implement a real-time system that could forecast ED arrivals on 1,2,3 hours horizon. Once measured the accuracy using the MAPE metric, it was noticed that great improvement happened when time of notification was incorporated into the model, especially on a one-hour horizon. Therefore, time of patient notification must be available for this model to be successful.

According to the studies mentioned earlier, it can be said that the existing studies have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established. Some of the studies said that the accuracy of hourly accident and emergency department forecasting model is low compared to other longer forecasting intervals like daily forecast [@boyle2012predicting; @hertzum2017forecasting]. However, some studies mentioned that the accuracy of ED hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].

There are few limitations in the literature which encourage us to undertake this research and propose examine different forecasting approaches to deal with them. These limitations are summarised as follows : (i) Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy. There is a lack of studies presenting probabilistic forecast of hourly ED arrivals that better represent uncertainty of future admissions, providing an holistic picture of future demand for a planner; (iii) most studies are not reproducible, as it is almost impossible to reproduce forecasting models and results from the studies; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e. there is no benchmark method nor is forecast accuracy reported.

@@

## data {#data}

Data used in the study comprised counts of patients’ arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital. We aggregated the patients’ arrival times to obtain hourly arrivals, which is used for empirical evaluation in this study.

Figure \@ref(fig:hourly-plot-ridge) illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy but it reveals some systematic structures.

It is clear that the number of arrivals has a sub-daily structure. The ED arrivals decreases between mid-night and early morning and then increases until the evening and then decreases again. It is also clear that ED service gets systematically more visits on Monday between 8 a.m. and 5 p.m. Moreover, The number of arrivals around mid-night is  slightly higher for Saturday and Sunday.

@@

Figure \@ref(fig:hourly-plot) also highlights that there are many ouliers for almost every hour of the day that may affec the accuracy of forecasting methods, some of these outliers are might be related to holidays and special events. It is also clear that arrivals are less volatile between mid-night and early morning.

We propose a forecasting model that accounts for the systematic structure of the data.

```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width= "70%"}

## Benchmarks {#benchmarks}

### Naive/Climatology

Empirical distribution for same hour-of-day and day-of-week as target time.

### Multiple Regression

Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the A&E admittance). The classical linear regression model is formulated as:

\begin{equation}
  \mathbf{y}_t = \mathbf{x}_t' \boldsymbol{\beta} + \epsilon_t ,
(\#eq:linearRegression)
\end{equation}

where $\mathbf{x}_t$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of parameters and $\epsilon_t$ is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance. However, in the context of healthcare and A&E admittance, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative. So the linear regression model should be substituted by some other model. One of the models that is frequently used in practice is the Poisson regression, which can be summarised as:

\begin{equation}
  \mathbf{y}_t \sim \mathrm{Poisson} \left( \exp \left( \mathbf{x}_t' \boldsymbol{\beta} \right) \right).
(\#eq:PoissonRegression)
\end{equation}

The exponent in \@ref(eq:PoissonRegression) is needed in order to make sure that the parameter of Poisson distribution is always positive. This model can be estimated via maximisation of the likelihood function based on Poisson mass function. When it comes to selecting explanatory variables for the model, there is no one correct answer and the decision needs to be done based on each specific case. In our experiment, we will only include dummy variables, capturing a variety of calendar events:

@@

The variables (1) - (3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events. Finally (5) is needed in order to capture the potential phenomenon of change in admittance after the holiday (e.g. people might try not to go to hospital on Christmas eve and thus will go the next day). This model assumes that all these effects are deterministic and do not change over time, but the exponentiation in \@ref(eq:PoissonRegression) introduces an interaction effect between dummy variables, so that the 3pm on Monday in January will be different from 3pm on Monday in July, although the parameters for hour of day and day of week are fixed and do not change over time. We use `alm()` function from `greybox` v0.7.0 package [@Svetunkov2021Greybox] for R [@RTeam2021] for the experiments and denote this model as "Poisson Regression".

### ETS - Exponential Smoothing model

@Hyndman2008b developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal component (thus *ETS*). Given the popularity of ETS model, we decided to include the basic ETS(A,N,A) model with the seasonal component with frequency 24 (hour of day) as a benchmark. This was done using `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and denote as *ETS*. This model does not capture the day of week or week of year effects, does not include explanatory variables, but its seasonal component and level change over time.

### Prophet

Prophet is a forecasting procedure created by Facebook [@taylor2018forecasting] that accounts for multiple seasonality, piecewise trend and holiday effects. Prophet is robust to missing data and shifts in the trend, and typically handles outliers well. Prophet works well on daily data seen in Facebook. It is popular and automated, making it easy to learn for beginners. The implementation may be less flexible than other methods.
The model is incorporated using corresponding implementation of the Fable package in R. We use the `prophet()` function in the fable package to generate hourly forecasts [@fable2020]. Note that but the input data is assigned with an hourly and daily seasonality.

### TBATS

@de2011forecasting proposed a model to deals with time series exhibiting multiple complex seasonalities. TBATS includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonality terms. The later one not only gives the model more flexibility to deal with complex seasonality but also reduces the parameters of model when the frequencies of seaonalities are high. We fit a TBATS model to our daily time series using the `tbats()` function in the forecast package of R [@forecastpackage2020].

# Model building {#model}

## ADAM: multiple seasonal iETSX

@SvetunkovAdam2021 proposed a framework for dynamic models called **ADAM** - Augmented Dynamic Adaptive Model. This framework includes ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting multiple frequencies, non-normal distributions and intermittent demand [@Svetunkov2019a]. Based on this framework, we use Gamma distribution for ETS(M,N,M) model with frequencies 24 (hour of day) and 168 (hour of week), adding dummy variables for week of year, holidays and lagged holidays. Given that the data exhibits zeroes, we use the direct probability of ETS(M,N,N) model developed by @Svetunkov2019a to treat those values. This model can be formulated as a set of the following equations:

\begin{equation}
	\begin{aligned}
	  & y_t = o_t z_t \\
		& \log z_t = \log l_{t-1} + \log s_{1,t-24} + \log s_{2,t-168} + \mathbf{x}_t' \boldsymbol{\beta} + \log \left(1 + \epsilon_{t} \right) \\
		& \log l_{t} = \log l_{t-1} + \log( 1  + \alpha \epsilon_{t}) \\ 
		& \log s_{1,t} = \log s_{1,t-m} + \log( 1  + \gamma_1 \epsilon_{t}) \\
		& \log s_{2,t} = \log s_{2,t-m} + \log( 1  + \gamma_2 \epsilon_{t}) \\
		& o_t \sim \text{Bernoulli} \left(\mu_{a,t} \right) \\
		& a_t = l_{a,t-1} \left(1 + \epsilon_{a,t} \right) \\
		& l_{a,t} = l_{a,t-1}( 1  + \alpha_{a} \epsilon_{a,t}) \\
		& \mu_{a,t} = \min(l_{a,t-1}, 1)
	\end{aligned} ,
	(\#eq:ADAMModel)
\end{equation}

where $\alpha$, $\beta$, $\gamma_1$, $\gamma_2$ and $\alpha_a$ are the smoothing parameters, defining how adaptive the components of the model should be, $l_t$ is the level component for the demand sizes, $s_{1,t}$ and $s_{2,t}$ are the seasonal components, $\boldsymbol{\beta}$ is the vector of parameters for the explanatory variables, $o_t$ is the binary variable, which is equal to one, when demand occurs and to zero otherwise, $l_{a,t-1}$ is the level component for the occurrence part of the model, and $\left(1+\epsilon_t \right) \sim \mathcal{\Gamma}(s^{-1}, s)$, where $s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2$ is the scale of the distribution. Finally, $a_t$ is an unobservable series, underlying the occurrence part of the model and $(1 + \epsilon_{a,t})$ is an unobservable error term for $a_t$. We expect this model to perform better than Poisson regression, because it has dynamic parts (level and seasonal components), but also takes external information into account. Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it. If integer-valued quantiles are needed, then rounding up can be done for them. This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM".

## GAMLSS

If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters. Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables. This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found. In practice, this means experimenting with various distributions and evaluating their suitability using available training data.
Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution. In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters. In a GAMLSS framework @Rigby2005 the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via

If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters. Generalised additive models for location, scale and shape (*GAMLSS*) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables. This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found. In practice, this means experimenting with various distributions and evaluating their suitability using available training data.

Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution. In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters. In a GAMLSS framework [@Rigby2005] the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via

@@

where $g_j$ is a monotonic `link` function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$. If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third element of $\bm x_t$. Each $f_{j,i}$ is a smooth function, constructed as

@@

As our data are counts, the natural starting point is the Poisson distribution, given by

\begin{equation}
  F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor}
  (\#eq:poissonreg)
\end{equation}

where we consider an additive model for $\lambda_t$ of the form

@@

\begin{equation}
  F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
 (\#eq:truncatedn)
\end{equation}

@@

## Forecast performance evaluation {#accuracy}

In order to assess performance of models, we track quantiles (5th, 10th, etc up to 95th quantile) and conditional expectations for 48 steps ahead for each model. The forecasts are produced every 12 hours for the holdout of 365 days in a rolling origin fashion [@Tashman2000], resulting in 727 origins. Based on these values, several error measures are calculated to evaluate the performance of models in terms of specific quantiles and in terms of expectation. The latter is measured via Root Mean Squared Error (RMSE):

\begin{equation}
  \mathrm{RMSE} = \sqrt{\frac{1}{n} \sum_{j=1}^n e_j^2} .
  (\#eq:RMSE)
\end{equation}

The objective of density forecasts it to be as sharp as possible while remaining reliable/calibrated [@Gneiting2007a]. A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, witch is valuable to decision makers provided the forecast is calibrated.

The Pinball Score is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score. It rewards sharpness and penalises mis-calibration, so measures all-round performance, however, calibration should still be verified separately. Furthermore, 
The Pinball Score for an individual quantile matches the loss function minimised in quantile regression model estimation. The Pinball Score is given by

\begin{equation}
    \text{Pinball} = 
    \frac{1}{T|\mathcal{A}|} \sum_{\alpha \in \mathcal{A}} \sum_{t=1}^T
 \left(q_{\alpha,t} - y_{t} \right)
 \left(\mathbb{1}(y_{t}\leq_{\alpha,t})-\alpha \right)
 (\#eq:pinball)
\end{equation}

where $\mathcal{A} = {0.05,0.1,...,0.95}$ is the set of quantiles being estimated.

>>>>

@morzuch2006forecasting used the Unobserved Components Model (UCM), in which each component of the time series is separately modeled as stochastic. Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for an horizon of 168 hours. The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles. The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.

@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services. Hourly data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future. The authors presented the prediction intervals with $50\%$ and $90\%$ confidence levels for the number of hourly arrivals under the Poisson assumption. They showed that the most important predictor was hour of the day and AR(1).
<!-- !!!! IS: I don't understand what is meant by the next sentence !!!! -->
<!-- However, these findings are limited to the short number of observations (only one year of historical data). -->

@schweigler2009forecasting conducted an investigation on whether time series methods could accurately generate short-term forecasts of ED bed occupancy.  A year-long dataset of hourly ED bed occupancy was collected from three facilities. For each facility, the authors implemented an hourly historical average model, SARIMA model and sinusoidal model with autocorrelated error. The historical average model was based on the mean occupancy for each site, for each hour of the day; while the sinusoidal model was based on 4 parameters: an AR term, a sine coefficient, a cosine coefficient and an intercept. They evaluated the forecast accuracy of four and twelve hours forecast horizon using RMSE and they found that both SARIMA and the sinusoidal models outperformed the historical average (for example, at site 2, the two models improved by $33\%$ the 12-hour forecasts generated by historical average).

@kim2014predicting compared different univariate and multivariate time series forecasting techniques to forecast patient volume for a Hospital Medicine programme. The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, GARCH (generalized autoregressive conditional heteroskedasticity method) and vector autoregressive (VAR) models to forecast for 4 hours and 24 hours ahead. They used MAPE to report the forecast accuracy. ARIMA model outperformed all the other models.

@gijo2016sarima applied a time series model to forecast the daily and hourly call volume at all centre handling emergency ambulance services. Since historical data showed seasonality, SARIMA models were used. Regarding the daily model, the authors used a SARIMA model, which, however, resulted in the forecast error that was significantly higher when the lead time exceeded 8 days. On the other hand, the SARIMA model was found to fit well the data both for shorter and longer lead times.

@asheim2019real developed a Poisson time-series regression model with continuous day-of-week and week-of-year effects to implement a real-time system that could forecast ED arrivals on 1, 2, 3 hours ahead. Measuring the accuracy using the MAPE, @asheim2019real noticed that great improvement happened when time of notification was incorporated into the model, especially in one-hour horizon.

According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established. Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data forecasting model is low compared to the higher forecasting intervals, like daily [@boyle2012predicting; @hertzum2017forecasting]. However, some studies mentioned that the accuracy of ED hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].

There are few limitations in the literature which encourage us to undertake this research and examine different forecasting approaches. These limitations are summarised as follows:

(i) Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time;
(ii) Almost all research studies produce point forecasts and consequently report only point forecast accuracy. There is a lack of studies presenting probabilistic forecasts of hourly ED arrivals that better represent uncertainty of future admissions, providing a holistic picture of future demand for a planner;
(iii) most studies are not reproducible, as it is almost impossible to reapply the approaches without the helpf of the authors of those papers;
(iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and
(v) some studies in this area lack a rigorous experimental design, i.e. there is no benchmark method or forecast accuracy is not reported.

@@

## Preliminary analysis {#data}

Data used in this study comprises counts of patients’ arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital. We aggregated the patients’ arrival times to obtain hourly arrivals, which is used for empirical evaluation in this study.

Figure \@ref(fig:hourly-plot-ridge) illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy, it reveals some systematic patterns.

@@

It is clear that the number of arrivals has a sub-daily structure, similar to the one summarised by @hertzum2017forecasting. The ED arrivals decrease between mid-night and early morning and then increase until the evening, decreasing after that again. It is also clear that ED service gets systematically more visits on Mondays between 8 a.m. and 5 p.m. Moreover, The number of arrivals around mid-night is  slightly higher for Saturday and Sunday.

Figure \@ref(fig:hourly-plot-ridge) also highlights that there are many ouliers for almost every hour of the day that may affect the accuracy of forecasting methods. Some of these outliers might be related to holidays and special events. It is also clear that arrivals are less volatile between mid-night and early morning.

**IS: We need to plot day of week and week of year seasonalplots here and discuss. This will show multiple seasonality nature of the data.**

Based on this simple analysis, we should consider model that could take the following into account:

- Hour of day, day of week, and week of year seasonal patterns,
- Calendar events, such as holidays,
- Lags of calendar events to accommodate the potential changes in demand the next day after a holiday,
- Potentially changing level of data,
- Other events, such as football matches in the area.

We propose several forecasting models that account for the structures outlined above.

<!-- {r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width= "70%"} -->

# Model building {#model}
<!-- ## Benchmarks {#benchmarks} -->

## Naive/Climatology {#climatology}
We start with one of the simplest forecasting approaches used in practice - the one that assumes that in the next few hours, everything will be the same as in the similar hours of a similar day in the past. This is called "Naïve". In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of demand is used to forecast the future demand distribution. **We need references here...**

<!-- Empirical distribution for same hour-of-day and day-of-week as target time. -->

## Multiple Regression
Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the A&E admittance). The classical linear regression model is formulated as:

\begin{equation}
  {y}_t = \mathbf{x}_t' \boldsymbol{\beta} + \epsilon_t ,
(\#eq:linearRegression)
\end{equation}

where $\mathbf{x}_t$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of parameters, $\epsilon_t$ is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance, and $t$ is the time index. However, in the context of healthcare and A&E admittance, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative. So the linear regression model should be substituted by some other model. One of the models that is frequently used in practice is the Poisson regression [see for example, @mccarthy2008challenge], which can be summarised as:

\begin{equation}
  \begin{aligned}
    & {y}_t \sim \mathrm{Poisson} \left( \lambda_t \right) \\
    & \log \lambda_t = \mathbf{x}_t' \boldsymbol{\beta}
  \end{aligned}
(\#eq:PoissonRegression)
\end{equation}

The logarithm in \@ref(eq:PoissonRegression) is needed in order to make sure that the parameter of Poisson distribution is always positive. This model can be estimated via maximisation of the likelihood function based on Poisson mass function [reference needed]. When it comes to selecting explanatory variables for the model, there is no single correct answer, and the decision needs to be done for each specific case. In our experiment, we will only include dummy variables, capturing a variety of calendar events:

@@

The variables (1) - (3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events. Finally (5) is needed in order to capture the potential phenomenon of change in admittance after the holiday (e.g. people might try not to go to hospital on Christmas eve and thus will go the next day). This model assumes that all these effects are deterministic and do not change over time, but the exponentiation in \@ref(eq:PoissonRegression) introduces an interaction effect between dummy variables, so that the 3pm on Monday in January will be different from 3pm on Monday in July, although the parameters for hour of day and day of week are fixed and do not change over time. We use `alm()` function from `greybox` package [@Svetunkov2021Greybox] for R [@RTeam2021] for the experiments and denote this model as "Poisson Regression".

## ETS - Exponential Smoothing model
@Hyndman2008b developed a state space approach for exponential smoothing models, according to which the model can have a set of components, including different types of Error, Trend and Seasonal component (thus *ETS*). Given the popularity of ETS model in forecasting community, we decided to include the basic ETS(A,N,A) model with the seasonal component with frequency 24 (hour of day) as a benchmark. This was done using `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and denote as *ETS*. This model does not capture the day of week or week of year effects, does not include explanatory variables, but its seasonal component and level change over time. This model is included as a benchmark, only to see how the other models perform in comparison with it.

## Prophet
Prophet is a forecasting procedure created by Facebook [@taylor2018forecasting] that accounts for multiple seasonality, piecewise trend and holiday effects. Prophet is robust to missing data and shifts in the trend, and typically handles outliers well. Prophet works well on daily data seen in Facebook. It is robust and automated, making it easy to learn for beginners. The implementation may be less flexible than other methods. The model itself relies on Multiple Source of Error state space model, originally proposed by @Kalman1960. The model is incorporated using corresponding implementation of the Fable package in R. We use the `prophet()` function from the `fable` package [@fable2020]. Note that the input data is assigned with an hourly and daily seasonality.

## TBATS
@de2011forecasting proposed a model to deal with time series exhibiting multiple complex seasonalities called "TBATS". It includes a Box-Cox Transformation, ARMA model for residuals and a trigonometric expression of seasonal terms. The latter one not only gives the model more flexibility to deal with fractional seasonality but also reduces the parameters of model when the frequencies of seasonalities are high. We fit a TBATS model using the `tbats()` function from the `forecast` package in R [@forecastpackage2020].

## ADAM: multiple seasonal iETSX
@SvetunkovAdam2021 proposed a framework for dynamic models called **ADAM** - Augmented Dynamic Adaptive Model. This framework encompasses ARIMA [@Box1976], ETS [@Hyndman2008b] and regression, supporting multiple frequencies, non-normal distributions and intermittent demand [@Svetunkov2019a]. Based on this framework, we use Gamma distribution for ETS(M,N,M) model with frequencies 24 (hour of day) and 168 (hour of week), adding dummy variables for week of year, holidays and lagged holidays. This way we update the hour of day and day of week seasonal indices, keeping the week of year one fixed, thus reducing the number of estimated parameters. Given that the data exhibits randomly occurring zeroes, we use the direct probability model developed by @Svetunkov2019a to treat those values. This model can be formulated as a set of the following equations:

\begin{equation}
	\begin{aligned}
	  & y_t = o_t z_t \\
		& \log z_t = \log l_{t-1} + \log s_{1,t-24} + \log s_{2,t-168} + \mathbf{x}_t' \boldsymbol{\beta} + \log \left(1 + \epsilon_{t} \right) \\
		& \log l_{t} = \log l_{t-1} + \log( 1  + \alpha \epsilon_{t}) \\ 
		& \log s_{1,t} = \log s_{1,t-m} + \log( 1  + \gamma_1 \epsilon_{t}) \\
		& \log s_{2,t} = \log s_{2,t-m} + \log( 1  + \gamma_2 \epsilon_{t}) \\
		& o_t \sim \text{Bernoulli} \left(\mu_{a,t} \right) \\
		& a_t = l_{a,t-1} \left(1 + \epsilon_{a,t} \right) \\
		& l_{a,t} = l_{a,t-1}( 1  + \alpha_{a} \epsilon_{a,t}) \\
		& \mu_{a,t} = \min(l_{a,t-1}, 1)
	\end{aligned} ,
	(\#eq:ADAMModel)
\end{equation}
where $\alpha$, $\beta$, $\gamma_1$, $\gamma_2$ and $\alpha_a$ are the smoothing parameters, defining how adaptive the components of the model should be, $l_t$ is the level component for the demand sizes, $s_{1,t}$ and $s_{2,t}$ are the seasonal components, $\boldsymbol{\beta}$ is the vector of parameters for the explanatory variables, $o_t$ is the binary variable, which is equal to one, when demand occurs and to zero otherwise, $l_{a,t-1}$ is the level component for the occurrence part of the model, and $\left(1+\epsilon_t \right) \sim \Gamma(s^{-1}, s)$, where $s=\frac{1}{T} \sum_{t=1}^{T} e_{t}^2$ is the scale of the distribution. Finally, $a_t$ is an unobservable series, underlying the occurrence part of the model and $(1 + \epsilon_{a,t})$ is an unobservable error term for $a_t$. @Svetunkov2019a discuss how to estimate such model. We expect this model to perform on par with the Poisson regression, potentially outperforming it in some instances, due to the dynamic parts of the model (level and seasonal components). Although the data is integer-valued, we expect that Gamma distribution will be a good approximation for it. If integer-valued quantiles are needed, then rounding up can be done for them (see \@ref(quantiles_ceiling) for the explanation). This model is implemented in `adam()` function from `smooth` package [@Svetunkov2021Smooth] for R and is denoted in our experiment as "ADAM-iETSX".

## GAMLSS
If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters. Generalised Additive Models for Location, Scale and Shape (GAMLSS) are distributional regression models where the parameters may be modelled as additive functions of explanatory variables. This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found. In practice, this means experimenting with various distributions and evaluating their suitability using available training data.

Let $F_t(y_t)$ be a predictive cumulative probability distribution of $y_t$. In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters. In a GAMLSS framework of @Rigby2005 the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via:

<!-- If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters. Generalised additive models for location, scale and shape (*GAMLSS*) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables. This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found. In practice, this means experimenting with various distributions and evaluating their suitability using available training data. -->

<!-- Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution. In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters. In a GAMLSS framework [@Rigby2005] the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via -->

@@

where $g_j$ is a monotonic `link` function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$ **is ...**. If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third elements of $\bm x_t$. Each $f_{j,i}$ is a smooth function, constructed as

@@

As our data are counts, the natural starting point is the Poisson distribution with an additive model for $\log \lambda_t$ of the form:

<!-- !!!! IS: I'm not sure that we need CDFs and PMFs, because they are well known. !!!! -->

<!-- \begin{equation} -->
<!--   F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor} -->
<!--   (\#eq:poissonreg) -->
<!-- \end{equation} -->

<!-- where we consider an additive model for $\lambda_t$ of the form -->

@@

<!-- \begin{equation} -->
<!--   F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)} -->
<!--  (\#eq:truncatedn) -->
<!-- \end{equation} -->

@@

# Forecast performance evaluation {#accuracy}

In order to assess performance of models, we track quantiles (5th, 10th, etc up to 95th quantile) and conditional expectations for 48 steps ahead for each model. We focus on 48 hours ahead, because this is the operational horizon in the hospital, for which it is possible to make short-term changes in the shifts for nurses and doctors. The forecasts are produced every 12 hours for the holdout of 365 days in a rolling origin fashion [@Tashman2000], resulting in 727 origins. Based on these values, several error measures are calculated to evaluate the performance of models in terms of specific quantiles and in terms of expectation. The latter is measured via Root Mean Squared Error (RMSE):

\begin{equation}
  \mathrm{RMSE} = \sqrt{\frac{1}{h} \sum_{j=1}^h e_{t+j}^2} ,
  (\#eq:RMSE)
\end{equation}
where $h$ is the forecast horizon and $e_{t+j}$ is the point forecast error $j$ steps ahead.

The objective of density forecasts is to be as sharp as possible while remaining reliable/calibrated [@Gneiting2007a]. A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision makers provided the forecast is calibrated. Calibration, also called *reliability*, is the property that forecast probabilities match the observed frequency of realisations. If a forecast is calibrated, then, for example, $20\%$ of observations should fall below the $\alpha=0.2$ predictive quantile (with some tolerance based on the finite sample size). This property is necessary for forecast probabilities to be used in quantitative decision-making. Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, $\alpha$, against observed frequency mean ($\mathbb{1}(y_{t}\leq q_{\alpha,t})$
<!-- IS: Check the \mathbb thingy - it didn't work out in pdf -->
). We use several scores to assess the quantile performance of models.
